---
title: "Fitting the EWMV-SLC Model to BART Data with cmdstanr"
subtitle: "An annotated end-to-end tutorial: data preparation, multithreaded MCMC sampling, convergence diagnostics, and model comparison"
author: "Toys4EWMV-SLC Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    df-print: paged
    code-line-numbers: true
execute:
  echo: true
  warning: false
  message: false
bibliography: references.bib
---

# Introduction

This document is an expanded, fully annotated version of `_scripts/cmdstanr.R`. It walks through the complete workflow of fitting the **EWMV-SLC** model (exponential-weight mean–variance with **diminishing loss sensitivity**) to **BART** (Balloon Analogue Risk Task) behavioral data using the [`cmdstanr`](https://mc-stan.org/cmdstanr/) R interface.

The EWMV-SLC model implemented here was developed to capture *diminishing
loss sensitivity* during risky decision-making in gambling disorder
[@wei2026diminishing]. The model code is adapted from the **EWMV** model of **Park et al. (2021)** [@park2021exponential] and the **hBayesDM** project of **Ahn et al. (2017)** [@ahn2017hbayesdm]. Our contribution is a `cmdstanr`-based re-implementation that adds **within-chain parallelism** (CmdStan `reduce_sum` + `threads_per_chain`): chains run in parallel *and* each chain splits the per-subject likelihood across multiple CPU threads, making full use of multi-core CPUs and greatly improving fitting speed and efficiency.

Compared with the classical `rstan` workflow (see `_scripts/bart_fit.R`), `cmdstanr` has several practical advantages:

- It calls the **CmdStan C++ executable** directly, so model code is compiled once with a modern C++ toolchain and reused across sessions.
- It supports **within-chain threading** (`threads_per_chain`) together with Stan's `reduce_sum`, which is exactly what `bart_ewmv-slc.stan` needs to spread the per-subject likelihood across CPU cores.
- It ships with `posterior` + `bayesplot` integration, and the fitted object can be converted back to an `rstan` `stanfit` when legacy `rstan` tooling is required.

The pipeline covered here is:

1. Load the data and helper functions.
2. Set the sampling settings and model constants.
3. Preprocess the long-format trial data into the matrices the Stan model expects.
4. Compile the Stan model with multithreading support.
5. Run MCMC sampling with parallel chains *and* per-chain threads.
6. Inspect the results (summaries, draws, diagnostic summaries).
7. Check convergence with trace plots and R-hat / ESS diagnostics.
8. Persist the fit to disk (`.rds`) and convert it to an `rstan` object.
9. Compare models with leave-one-out cross-validation (LOO-CV / WAIC).

---

# Getting started: clone the repository

All commands in this tutorial run from the **repository root** (the folder with
`_data/`, `_scripts/`, `cmdstanr.qmd`, `README.md`).

## Command line basics (for beginners)

All commands here are typed in a **terminal**: a text-only window where you
type a command and press **Enter** to run it. Not sure how to open one?

### Windows

1. Press **`Win + R`**, type `cmd`, press Enter, or open **Windows
   PowerShell** from the Start menu (search "powershell").
2. You will see a prompt like `C:\Users\yourname>`. Type commands after it and
   press Enter. **Paste** commands with right-click or `Ctrl + V`.

| Command | What it does |
|---------|--------------|
| `dir` | List the files and folders in the current folder |
| `cd folder` | Enter a folder (e.g. `cd EWMV-SLC_Tutorial`) |
| `cd ..` | Go up one level |
| `cls` | Clear the screen |
| `%CD%` | The current folder path (what the Docker commands use) |

### Which Windows shell should I use?

Windows offers several terminal programs. They *look* the same but have
different syntax. The key difference here is **how they write the
current folder path** in the Docker commands:

| Shell | How to open | Path syntax | Notes |
|-------|-------------|-------------|-------|
| CMD (Command Prompt) | `Win + R` → `cmd` | `%CD%` | Classic; `dir` to list files; line continuation with `^` |
| PowerShell | Start menu → search "PowerShell" | `${PWD}` | Default modern shell; `dir` works too; line continuation with `` ` `` |
| Git Bash | Start menu → search "Git Bash" (installed with Git) | `$(pwd)` | Simulates a Linux environment (`ls`, `cd`); behaves like macOS/Linux |
| Windows Terminal (optional) | Microsoft Store app | any of the above | A modern tabbed window that can run CMD/PowerShell/Git Bash/WSL |

**Rule of thumb:** if you followed the Git option above, use **Git Bash**; then every command in this tutorial works verbatim (same as macOS/Linux).
If you use CMD or PowerShell, swap `$(pwd)` for `%CD%` or `${PWD}` accordingly.

### macOS / Linux

1. **macOS**: press **`Cmd + Space`**, type "Terminal", press Enter (or Finder
   → Applications → Utilities → Terminal).
2. **Linux**: press `Ctrl + Alt + T` (most distributions) or open the
   "Terminal" app. **Paste** commands with `Cmd + V` (macOS) or `Ctrl + V`
   (Linux); on older macOS versions right-click → Paste instead.

`cd` without a path, `cd ~`, or simply typing `cd` returns you to your home
folder. macOS/Linux use the same `cd`/`ls` style; Windows uses `dir`/`cd`
(with `cd /d C:\path` in CMD if you need to switch drives).

| Command | What it does |
|---------|--------------|
| `ls` | List the files and folders in the current folder |
| `cd folder` | Enter a folder (e.g. `cd EWMV-SLC_Tutorial`) |
| `cd ..` | Go up one level |
| `pwd` | Print the current folder path |
| `$(pwd)` | The current folder path (what the Docker commands use) |

### Always work from the repository root

Before running any command, `cd` into the project folder you created below and
confirm you see `_data`, `_scripts`, `cmdstanr.qmd`, `README.md` with `dir`/`ls`.
If a command fails, the most likely cause is being in the wrong folder.

### Common errors (and how to fix them)

| Error message | What it usually means | Fix |
|---------------|-----------------------|-----|
| `'docker' is not recognized` / `command not found` | Docker not installed or not started | Install Docker (links above), restart the terminal, start Docker Desktop |
| `cannot connect to the Docker daemon` | Docker Desktop is not running | Start Docker Desktop from the Start menu / Launchpad |
| `invalid reference format` | Bad path syntax (usually `$(pwd)` in PowerShell) | Use `"${PWD}:/root/stan"` in PowerShell, `%CD%` in CMD, `$(pwd)` in Bash/Zsh |
| `exec: "-v": executable file not found` | `-v` was typed after the image name | Put all options (`-v`, `-p`, `-e`) **before** the image name |
| `port is already allocated` | Port 8787 is in use | `docker ps` → `docker stop <name>`, or use `-p 8788:8787` and open `localhost:8788` |
| `the container name is already in use` | A container with the same name exists | `docker rm -f ewmv-rstudio` and run again |
| `unable to find image ... locally` / pull timeout | Network problem downloading the image | Retry; configure a registry mirror or proxy for Docker |
| `The system cannot find the path specified` / `No such file or directory` | Wrong current folder or wrong mount path | `cd` to the repository root; re-check with `dir`/`ls` |
| Browser shows a blank page | Container not running | `docker ps`; the container status must be `Up` |

If you still have a problem, copy the full error text into a search engine or
an AI assistant, and mention you are using `koalalive/stan4cogneuro:1.0.2`.

## Option A: git clone (recommended)

Install Git if you do not have it yet:

- **Windows**: <https://gitforwindows.org/>
- **macOS**: `xcode-select --install` or <https://git-scm.com/download/mac>
- **Linux**: your package manager, e.g. `sudo apt install git`

```bash
# 1. Clone the repository
git clone https://github.com/Koalalive/EWMV-SLC_Tutorial.git

# 2. Enter the project folder
cd EWMV-SLC_Tutorial

# 3. Verify you are in the right place (you should see the folders below)
ls
# _data  _scripts  cmdstanr.qmd  README.md  AGENTS.md  ...
```

## Option B: download as ZIP (no Git needed)

Open <https://github.com/Koalalive/EWMV-SLC_Tutorial> in your browser, click
**Code → Download ZIP**, unzip the archive, and open a terminal inside the
unzipped folder. **Run every command below from that folder**; it is the
repository root used throughout this tutorial.

## What you should see

| Path | What it is |
|------|------------|
| `_data/bart_sample.csv` | Sample BART data |
| `_scripts/bart_ewmv-slc.stan` | The EWMV-SLC Stan model |
| `_scripts/cmdstanr.R` | The main fitting pipeline |
| `cmdstanr.qmd` | This tutorial (annotated source) |
| `README.md` | Human-facing documentation |

Next: [Prerequisites](#prerequisites) to get Docker ready (see the Docker
primer in the README if you are new to Docker).

---

# Prerequisites

## Recommended: use the Docker environment (zero setup)

The image `koalalive/stan4cogneuro:1.0.2` bundles R, CmdStan, `cmdstanr`,
`posterior`, `bayesplot`, `loo`, `tidyverse`, and `rstan`. With Docker you do
not need to install anything locally. The same pinned environment runs on any
computer, which makes the analysis easy to deploy and the results easy to
reproduce:

```bash
# Pull the environment
docker pull koalalive/stan4cogneuro:1.0.2

# Run the whole pipeline inside the container
# (PowerShell: ${PWD}, CMD: %CD%, Bash/Zsh: $(pwd))
# NOTE: R is not on the default PATH, use the full path below.
docker run --rm -v "$(pwd)":/root/stan -w /root/stan \
  koalalive/stan4cogneuro:1.0.2 \
  /root/miniconda3/envs/stan/bin/Rscript _scripts/cmdstanr.R
```

::: {.callout-tip}
## Working interactively (RStudio Server)

For an interactive IDE, mount the project to `/root/stan`, publish port 8787, and open `http://localhost:8787` in your browser (default user: `rstudio-server`, password `rstudio`). RStudio Server starts automatically. Pick your shell:

```bash
# Bash (macOS / Linux)
docker run -it --name ewmv-rstudio -p 8787:8787 \
  -v "$(pwd)":/root/stan \
  koalalive/stan4cogneuro:1.0.2
```

```zsh
# Zsh (macOS / Linux)
docker run -it --name ewmv-rstudio -p 8787:8787 \
  -v "$(pwd)":/root/stan \
  koalalive/stan4cogneuro:1.0.2
```

```powershell
# Windows PowerShell
docker run -it --name ewmv-rstudio -p 8787:8787 `
  -v "${PWD}:/root/stan" `
  koalalive/stan4cogneuro:1.0.2
```

```bat
:: Windows CMD
docker run -it --name ewmv-rstudio -p 8787:8787 -v "%CD%":/root/stan koalalive/stan4cogneuro:1.0.2
```
:::

::: {.callout-tip}
## Rendering this tutorial

The image does not ship the Quarto CLI, so `cmdstanr.qmd` is meant to be read as annotated source and its R chunks executed inside **RStudio Server** or any R session with the packages below.
:::

## Alternative: local installation

Install the packages used in this tutorial:

```r
install.packages(c("cmdstanr", "posterior", "bayesplot", "loo", "tidyverse", "rstan"))
```

::: {.callout-note}
## cmdstanr requires a C++ toolchain

Before `cmdstanr` can compile models you must have a working C++ toolchain:

- **macOS**: Xcode command line tools.
- **Windows**: [RTools](https://cran.r-project.org/bin/windows/Rtools/) matching your R version.
- **Linux**: a recent `g++` and `make`.

Then install the CmdStan backend itself:

```r
cmdstanr::install_cmdstan()
```
:::

## Project layout

| File | Purpose |
|------|---------|
| `_data/bart_sample.csv` | Sample BART data in long format (one row per trial) |
| `_scripts/bart_fit.R` | `bart_fit_preprocess()` (data reshaping) and the `rstan`-based `bart_fit_ewmvslc()` reference implementation |
| `_scripts/bart_diagnosis.R` | `bart_ewmvslc_diagnosis()` (R-hat + trace plots of the group means) |
| `_scripts/bart_ewmv-slc.stan` | The EWMV-SLC Stan model (threaded with `reduce_sum`) |
| `_scripts/cmdstanr.R` | The plain R script this tutorial expands |
| `cmdstanr.qmd` | This document |
| `_rds/` | Saved fits (`*.rds`) and CmdStan output CSVs (`*.csv`) |

---

# Step 0: Setup, libraries, and helper functions

Load every package used in this workflow and source the project's helper functions (`bart_fit_preprocess` reshapes the data; `bart_ewmvslc_diagnosis` collects R-hat values and trace plots).

```{r}
#| label: setup

library(cmdstanr)   # R interface to CmdStan (compilation, sampling, summaries)
library(posterior)  # working with draws objects (draws_array / draws_matrix)
library(bayesplot)  # MCMC visualization (trace plots, histograms, intervals)
library(loo)        # leave-one-out cross-validation / WAIC
library(tidyverse)  # data manipulation and plotting (dplyr, ggplot2, readr...)

# Load the helper functions from the repository scripts.
#   bart_fit_preprocess()  : long-format data -> Stan-ready arrays
#   bart_ewmvslc_diagnosis(): R-hat and trace plots (see _scripts/bart_diagnosis.R)
source("_scripts/bart_fit.R")
source("_scripts/bart_diagnosis.R")
```

---

# Step 1: The model in brief

The Stan model (`_scripts/bart_ewmv-slc.stan`) is a **hierarchical Bayesian
model** [@gelman2013bda3]: each subject's six parameters are drawn from
group-level (population) distributions, so information is shared across
subjects while individual differences are preserved. The population means
(`mu_phi` … `mu_zeta`) are the quantities of interest for group comparisons
(e.g. gambling disorder vs. controls).

The subject-level parameters are written in a *non-centered* (Matt trick)
parameterization [@papaspiliopoulos2007general], which greatly improves MCMC
efficiency by removing the dependence between the group-level mean and the
subject-level deviations:

| Parameter | Meaning (as implemented in the model) | Constraint |
|-----------|----------------------------------------|------------|
| `phi[j]` | Initial subjective explosion probability for subject `j` | (0, 1) |
| `eta[j]` | Learning rate updating the subjective explosion probability from trial feedback | (0, 1) |
| `rho[j]` | Mean–variance (risk) correction weight in the estimated utility | (-0.5, 0.5) |
| `tau[j]` | Decision temperature (inverse sensitivity) of the softmax / logit choice rule | (0, ∞) |
| `lambda[j]` | Loss aversion | (0, ∞) |
| `zeta[j]` | Diminishing loss-sensitivity exponent applied to the loss magnitude | (0, ∞) |

At each pump step the model compares the utility of **pumping again** (`u_pump`) with the utility of **stopping** (`u_stop`, fixed at 0):

$$u_{\text{pump}} = (1-p)\,u_{\text{gain}} - \lambda\, p\, u_{\text{loss}}^{\zeta} + \rho\, p\,(1-p)\,\big(u_{\text{gain}} + \lambda\, u_{\text{loss}}^{\zeta}\big)^{2}$$ {#eq-utility}

where $u_{\text{gain}} = 1$, $u_{\text{loss}} = l-1$ grows with the number of pumps, and $p$ is the current subjective explosion probability. The decision follows a logit-link Bernoulli rule,

$$\Pr(\text{pump}) = \operatorname{logit}^{-1}\big(\tau \cdot (u_{\text{pump}} - u_{\text{stop}})\big)$$

The subjective explosion probability $p$ is updated after each trial as a function of the observed outcome via the learning rate `eta[j]`.

::: {.callout-note}
## Threading

The likelihood is computed with Stan's `reduce_sum`, so the per-subject evaluation can be split across worker threads. For this to work the model **must** be compiled with `cpp_options = list(stan_threads = TRUE)` (see @sec-compile).
:::

---

# Step 2: Global settings

The naming follows the `rstan`-style settings used in `_scripts/cmdstanr.R`, then is mapped onto the `cmdstanr` arguments:

- `n.iter` = total iterations per chain,
- `n.burnin` = warmup iterations,
- `n.chain` = number of Markov chains,
- `Trials` = number of trials per subject (pad `T`),
- `Nmax` = maximum number of pumps (used to define `P = Nmax + 1`).

```{r}
#| label: settings

# --- Sampling settings -----------------------------------------------------
n.iter   <- 2000 # total iterations per chain (warmup + sampling)
n.burnin <- 1000 # number of burn-in (warmup) iterations
n.chain  <- 4    # number of Markov chains

# --- Data constants ---------------------------------------------------------
Trials <- 100 # number of trials per subject in the sample data
Nmax   <- 12  # maximum number of pumps allowed in a trial
```

::: {.callout-tip}
## Mapping to cmdstanr argument names

`cmdstanr` uses `iter_warmup` and `iter_sampling` separately. Here this means: `iter_warmup = n.burnin` and `iter_sampling = n.iter - n.burnin` (1000 draws of posterior samples per chain in this example).
:::

---

# Step 3: Load the data

The sample data are in a tidy long format with one row per trial:

- `subjID`: participant identifier,
- `pumps`: number of pumps on that trial,
- `explosion`: 1 if the balloon exploded (the last pump on this trial was an explosion), 0 otherwise.

```{r}
#| label: load-data

# Read the formatted trial-level data
bart_sample <- read_csv("_data/bart_sample.csv")

# Take a quick look at the structure and the first rows
str(bart_sample)
head(bart_sample, 10)

# How many subjects and trials do we have?
bart_sample |>
  group_by(subjID) |>
  summarise(n_trials = n(), .groups = "drop") |>
  print()
```

---

# Step 4: Preprocess the data for Stan

The Stan model expects **matrices** of shape `N x T`, not a long data frame. `bart_fit_preprocess()` converts the long format into the `data_list` that `model$sample()` consumes:

- `N`: number of subjects,
- `T`: maximum number of trials per subject (padded with 0s if trials differ),
- `Tsubj`: actual number of trials per subject,
- `P`: maximum number of pumps + 1 (`Nmax + 1`),
- `pumps`: `N x T` integer matrix,
- `explosion`: `N x T` integer matrix (0/1).

```{r}
#| label: preprocess

data_list <- bart_sample %>%
  bart_fit_preprocess(., Trials, Nmax) # reshape long data into Stan arrays

# Inspect the resulting data list
str(data_list)
```

The model then derives the `d[j, k, l]` indicator (1 = the participant pushed the pump at step `l` of trial `k`) inside the `transformed data` block:

```{r}
#| label: show-transformed-data
#| echo: false

cat("N =", data_list$N, " T =", data_list$T, " P =", data_list$P, "\n")
cat("Tsubj =", paste(head(data_list$Tsubj, 5), collapse = ", "), "...\n")
```

---

# Step 5: Compile the Stan model {#sec-compile}

Compiling translates the `.stan` file into a C++ binary. Because the model uses `reduce_sum`, we must enable `stan_threads = TRUE` so that `threads_per_chain` is honored later.

```{r}
#| label: compile-model
#| cache: true

# Model file (renamed from bart_ewmv-slc_wcp_new.stan to bart_ewmv-slc.stan)
model_file <- "_scripts/bart_ewmv-slc.stan"

# Compile the model with multithreading support
model <- cmdstan_model(model_file,
                       cpp_options = list(stan_threads = TRUE))
```

::: {.callout-note}
## First run takes a minute (or longer)

The **first** execution of `cmdstan_model()` triggers a C++ compilation that typically takes about one minute on a typical desktop. Subsequent runs reuse the cached executable and start immediately.
:::

---

# Step 6: Run the MCMC sampler {#sec-sample}

`CmdStanMCMC$sample()` runs the compiled executable. Two levels of parallelism are used:

- `parallel_chains = 4`: run the 4 chains *simultaneously* using 4 processes;
- `threads_per_chain = 3`: each chain further divides the per-subject likelihood across 3 threads (via `reduce_sum`), i.e. 12 cores in total.

```{r}
#| label: sample
#| cache: true

# Number of threads used within each chain (reduce_sum worker threads)
threads_per_chain <- 3

# Start timing (fitting may take several minutes depending on the CPU)
start_time <- Sys.time()

# Preprocess the data, then call cmdstan to do the sampling
bart_fit_ewmvslc_sample_cmdstan <- bart_sample %>%
  bart_fit_preprocess(., Trials, Nmax) %>%
  model$sample(
    data = .,
    chains = n.chain,
    parallel_chains = n.chain,            # chains in parallel
    threads_per_chain = threads_per_chain, # threads within each chain
    iter_warmup = n.burnin,               # warmup (burn-in) iterations
    iter_sampling = n.iter - n.burnin,    # posterior draws per chain
    seed = 1234,                          # reproducible random seed
    output_dir = "_rds"                   # keep CmdStan CSV output files on disk
  )

# Show the elapsed fitting time
print(Sys.time() - start_time)
```

::: {.callout-tip}
## Why `output_dir = "_rds"`?

By default `cmdstanr` writes the per-chain sampler output (CSV files) to a temporary directory. Setting `output_dir = "_rds"` keeps them in the project folder, which is useful for

- keeping the raw sampler output as evidence / for `read_stan_csv()` later,
- inspecting the CSV files directly, and
- recycling the fit after an `.rds` round-trip (see @sec-persist).
:::

---

# Step 7: Inspect the fitted object

A `CmdStanMCMC` object is rich in methods. The most commonly used ones:

- `$draws()`: extract the posterior draws (as `draws_array`),
- `$summary()`: a tidy table with `median`, `mad`, `rhat`, `ess_bulk`, `ess_tail`,
- `$diagnostic_summary()`: warmup + sampling diagnostics,
- `$output_files()`: paths to the per-chain CmdStan CSV output files.

```{r}
#| label: inspect-summary

# Posterior summaries for every parameter
summary_tbl <- bart_fit_ewmvslc_sample_cmdstan$summary()
print(summary_tbl)

# Raw draws object (iterations x chains x parameters)
draws_posterior <- bart_fit_ewmvslc_sample_cmdstan$draws()
print(draws_posterior)
```

```{r}
#| label: diagnostics-summary

# All sampler diagnostics of the fitted run
fit_diagnostics <- bart_fit_ewmvslc_sample_cmdstan$diagnostic_summary()
print(fit_diagnostics)
```

::: {.callout-note}
## What to look for

- `rhat` close to 1.00 (rule of thumb: `< 1.01`) suggests the chains converged.
- `ess_bulk` / `ess_tail` above a few hundred indicate sufficient effective posterior sample size for stable estimates.
- `num_divergent` / `num_max_treedepth` should be 0. If not, consider more warmup, a better prior, or reparameterization.
:::

---

# Step 8: Convergence diagnostics

The `bayesplot` interface works directly with the `draws` objects from `cmdstanr`. Trace plots let you eyeball whether the chains mix well and cover the same posterior region.

The **R-hat** diagnostic originates from Gelman & Rubin
[@gelman1992inference], who compared between- and within-chain variance of
multiple chains. The version implemented here (in `posterior`/`cmdstanr`) is
the **rank-normalized, split R-hat** of Vehtari et al. [@vehtari2021rank],
which is far less sensitive to non-normal posterior distributions and is the
currently recommended convergence diagnostic. Values close to 1.00 (rule of
thumb: below 1.01) indicate that the chains have converged; larger values
usually mean more warmup or iterations are needed.

```{r}
#| label: rhat-all
#| fig-cap: "R-hat values for every parameter"

# R-hat for every parameter; close to 1.00 means the chains converged
rhat_all <- posterior::rhat(draws_posterior)
print(rhat_all)
```

```{r}
#| label: rhat-check

# Quick check: flag any parameter with R-hat above 1.01
bad_rhat <- names(rhat_all)[rhat_all > 1.01]
if (length(bad_rhat) == 0) {
  cat("Convergence check: all parameters have R-hat <= 1.01.\n")
} else {
  cat("Parameters with R-hat > 1.01 (may need more iterations):",
      paste(bad_rhat, collapse = ", "), "\n")
}
```

```{r}
#| label: trace-group-means
#| fig-cap: "Trace plots of the group-level mean parameters"

# Trace plots for all group-level means (phi, eta, rho, tau, lambda, zeta)
draws_posterior %>%
  bayesplot::mcmc_trace(regex_pars = "mu_")
```

```{r}
#| label: trace-mu-zeta
#| fig-cap: "Trace plot of the group-level loss-sensitivity exponent mu_zeta"

# Same as the script: focus on mu_zeta, the loss-sensitivity exponent
draws_posterior %>% bayesplot::mcmc_trace(regex_pars = "mu_zeta")
```

```{r}
#| label: hist-group-means
#| fig-cap: "Posterior histograms of the group-level means"

# Marginal posterior distributions of the group means
draws_posterior %>%
  bayesplot::mcmc_hist(regex_pars = "mu_", binwidth = 0.05)
```

```{r}
#| label: intervals-group-means
#| fig-cap: "Posterior intervals of the group-level means"

# Median plus 50% and 95% credible intervals
draws_posterior %>%
  bayesplot::mcmc_intervals(regex_pars = "mu_")
```

---

# Step 9: Persist the fit and convert to an rstan object {#sec-persist}

## Save and reload

A `CmdStanMCMC` fit can be stored as an `.rds` file just like any R object:

```{r}
#| label: save-fit

saveRDS(bart_fit_ewmvslc_sample_cmdstan,
        file = "_rds/bart_fit_ewmvslc_sample_cmdstan.rds")

# Reload it in a fresh session
bart_fit_ewmvslc_sample_cmdstan <- readRDS("_rds/bart_fit_ewmvslc_sample_cmdstan.rds")
```

## Conversion to rstan (optional)

Some older ecosystem tools expect an **rstan `stanfit`** object. A `stanfit` can be rebuilt from the CmdStan **CSV output files** of the run, one CSV per chain,. This is the reason we kept `output_dir = "_rds"` in @sec-sample:

```{r}
#| label: convert-to-rstan
#| eval: false

library(rstan) # needed for read_stan_csv()

# IMPORTANT: read_stan_csv() reads the raw CmdStan CSV output files
# (fit$output_files()), NOT a draws object. The CSVs must still exist on disk.
stanfit <- rstan::read_stan_csv(bart_fit_ewmvslc_sample_cmdstan$output_files())

# The resulting object works with rstan's usual post-processing functions
print(stanfit)
```

::: {.callout-warning}
## Old script pitfall

The original `cmdstanr.R` passed `as_draws_df(samples_posterior)` to `read_stan_csv()`. That does **not** work: `read_stan_csv()` parses CSV files produced by CmdStan, not in-memory draws. Use `fit$output_files()` as above; or, if the CSVs were discarded, refit with `output_dir` set (see @sec-sample).
:::

---

# Step 10: Model comparison with LOO-CV

The pointwise log-likelihood (`log_lik` in the `generated quantities` block)
is what we need for leave-one-out cross-validation. The LOO-CV estimate of
the expected log pointwise predictive density, computed with Pareto-smoothed
importance sampling (PSIS-LOO) [@vehtari2017practical], is implemented in the
`loo` package:

```{r}
#| label: loo

# Extract the pointwise log-likelihood matrix (iterations x N observations)
log_lik_sims <- bart_fit_ewmvslc_sample_cmdstan$draws("log_lik")

# Compute the LOO-CV estimate of the expected log predictive density
bart_fit_loo <- loo(log_lik_sims)

# Summarize the LOO-CV result
print(bart_fit_loo)
```

::: {.callout-note}
## Reading the LOO output

- `elpd_loo`: estimated log pointwise predictive density (higher is better).
- `p_loo`: effective number of parameters (penalty for overfitting).
- `looic`: the LOO information criterion, `-2 * elpd_loo` (lower is better).
- **Pareto `k`** diagnostics: values below 0.5 are desirable; values above 0.7 flag influential observations that need special care (e.g. `kfold()`).

Once you have LOO results for more than one model (e.g. EWMV vs. EWMV-SLC), compare them with `loo_compare(list("EWMV-SLC" = fit_a_loo, "EWMV" = fit_b_loo))`.
:::

---

# Where to Go From Here

- **Tune `grainsize`** in `bart_ewmv-slc.stan` (currently 1). A larger value reduces scheduler overhead but can hurt load balancing; a common starting point is `N / (chains * cores)`.
- **Increase `iter_sampling`** if the summary shows small `ess_bulk`.
- **Compare against nested models** (e.g. a model without the `rho` term or a version with `zeta = 1`) using `loo_compare()`.
- **Swap in your own data**: keep the same `subjID`, `pumps`, `explosion` columns and adjust `Trials` / `Nmax` accordingly.
- **Replicate across simulated or multiple empirical datasets** to study the recovery and stability of the EWMV-SLC parameters.
- **Apply to clinical data**: the empirical application (gambling disorder vs. controls, group-level hypothesis tests on `mu_zeta` and `mu_lambda`) is described in the source paper [@wei2026diminishing].
## Acknowledgments

We sincerely thank **Gangliang Zhong**, **Yujie Bai**, **Lingjie Wei**,
**Aimin Zhao**, **Fei Chu**, and **Xiao Zuo** for their generous provision of
computational resources and testing support, which made this repository
possible.
## Software references

- `cmdstanr` R interface to CmdStan: <https://mc-stan.org/cmdstanr/>
- `posterior` package: <https://mc-stan.org/posterior/>
- `bayesplot` plotting functions: <https://mc-stan.org/bayesplot/>
- `loo` package (PSIS-LOO, Pareto k diagnostics): <https://mc-stan.org/loo/>
- Stan documentation (`reduce_sum`, threading): <https://mc-stan.org/docs/>
